Towards Compiler-Agnostic Performance in Finite-Difference Codes. (TBC)

### 2014-11-12 shallow_perf.tex
Beginnings of detailed discussion of results for 256 case

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
Originally written by Paul Schwartrauber of NCAR.  Organic development
- optimised for previous hardware.

@@

For completeness, we give the compiler flags used with each compiler
in Table~\ref{TABLE_compiler_flags}. These flags are particularly
important for the \psykal versions of the code. The performance of the
original version of the code is very much less sensitive to the
compiler options. This is because it consists of a single source file
and a single routine (excluding those for I/O).

@@

In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and yellow bars. The performance of the Ivy
Bridge system (red and purple bars) only drops to this level when the
domain size is increased to $1024^{2}$. At this point, the working set
no longer fits within cache and performance is dominated by the
bandwidth to main memory making it relatively insensitive to the
choice of compiler.

For problem sizes that fit within cache there is
little to choose between the performance obtained from the Intel and
Gnu compilers.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
\caption{Summary of the performance of the original shallow code on 
the range of compilers and hardware under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

@@

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
\hline
CPU:                       & Haswell & Haswell & Ivy Brige & Ivy Bridge & PowerPC A2 \\
\hline
Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
\hline
\%-slow-down c.f. original &  45.3   & 11.7   & 5.0   & 11.0  &        \\
\hline
\end{tabular}
\end{table*}

>>>>

\subsection{The `shallow' Program}
For this work we have used a benchmarking program that solves the
shallow-water equations on a bi-periodic plane following the
finite-difference scheme introduced by Sadourney~\cite{sadourny75}.
Originally written by Paul Swarztrauber of the National Center for
Atmospheric Research, US.  Organic development - optimised for
previous hardware. In describing our work, we term the version of the
shallow code obtained at the beginning of this work the `original'
version.

@@

In Table~\ref{TABLE_compiler_flags} we give the optimisation flags
used with each compiler. These flags are particularly important for
the \psykal versions of the code. The performance of the original
version of the code is very much less sensitive to the compiler
options. This is because it consists of a single source file
containing effectively a single routine (excluding those for I/O).

@@

In figure~\ref{FIG_orig_perf_summary} we plot the performance of the
original version of the shallow code for the range of compilers and
hardware considered here. This summary demonstrates the effect of the
larger last-level cache of the Intel Ivy Bridge system compared to
that of the Haswell system; note the significant drop in performance
in going from a problem size of $256^{2}$ to a problem size of
$512^{2}$ for the black and yellow bars. The performance of the Ivy
Bridge system (red and purple bars) only drops to this level when the
domain size is increased to $1024^{2}$. At this point, the working set
no longer fits within cache and performance is dominated by the
bandwidth to main memory, making it relatively insensitive to the
choice of compiler.

For problem sizes that fit within cache there is little to choose
between the performance obtained from the Intel and Gnu compilers on
the Intel Haswell CPU.

\begin{figure}[!t]
\centering
\includegraphics[width=2.8in]{orig_summary}
\caption{Summary of the performance of the original shallow code on 
the range of compilers and hardware under consideration.}
\label{FIG_orig_perf_summary}
\end{figure}

@@

In Table~\ref{TABLE_opt_breakdown} we show detailed timings of the
code after each of the transformations has been applied. For the Gnu
compiler, the vanilla \psykal version of shallow is some 20 times
slower than the original. Copying the code for each of the kernels
into the module from which they are called has the most dramatic
effect on the performance of the compiled code. None of the subsequent
transformations succeed in significantly improving the performance of
the compiled executable which is still some 45\% slower than the
original.
{\bf What is losing us this 45\%?}

\begin{table*}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{The time (s) taken to do 2000 steps of the $256^2$ case after each 
code transformation.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c|c|c}
\hline
Compiler:                  &  Gnu    & Intel & Cray & Intel & IBM bgxlf \\
\hline
CPU:                       & Haswell & Haswell & Ivy Bridge & Ivy Bridge & PowerPC A2 \\
\hline
Original                   &  1.49   & 1.41   & 1.24  & 1.57  &        \\
Vanilla \psykal            & 25.27   & 1.73   & 1.67  & 1.96  &        \\
Explicitly-sized arrays    & 25.28   & 1.72   & 1.39  & 1.93  & 113.74 \\
Kernels copied into module &  2.20   & 1.73   & 1.42  & 1.93  &  68.28 \\
Loops fused                &  2.18   & 1.58   & 1.47  & 1.76  &  71.91 \\
In-lined field-copy        &  2.17   & 1.58   & 1.46  & 1.75  &  83.36 \\
Fused field-copy           &  2.18   & 1.85   & 1.30  & 2.07  &  72.62 \\
\hline
\parbox{2.5cm}{\%-slow-down of best c.f. original} &  45.3   & 11.7   & 5.0   & 11.0  &        \\
\hline
\end{tabular}
\end{table*}

% 0.32/1.41 = 23% and 0.39/1.57 = 25%
Considering now the Intel compiler, Table~\ref{TABLE_opt_breakdown}
shows that even the vanilla \psykal version of shallow is only
23--25\% slower than the original version. The only code
transformation that produced a significant performance improvement for
this case was the fusion of the computational loops. Further
investigation of the slow-down caused by attempting to fuse the
field-copy operation with the preceeding time-smoothing loop revealed
that the compiler was unable to vectorise the loop after it had
in-lined the body of the time-smoothing kernel. Once this in-lining
was done by hand, the compiler was able to vectorise the loop and the
performance of the original version was recovered.

\begin{figure}
\begin{verbatim}
! Time smoothing
DO J=1,N+1
  DO I=1,M+1
    CALL time_smooth_code(i,j,ufld,unew,uold)
!    uold(i,j) = ufld(i,j) + alpha* &
!        (unew(i,j) - 2.*ufld(i,j) + uold(i,j))

    CALL time_smooth_code(i,j,vfld,vnew,vold)
!    vold(i,j) = vfld(i,j) + alpha* &
!        (vnew(i,j) - 2.*vfld(i,j) + vold(i,j))

    CALL time_smooth_code(i,j,pfld,pnew,pold)
!    pold(i,j) = pfld(i,j) + alpha* &
!        (pnew(i,j) - 2.*pfld(i,j) + pold(i,j))
  END DO
END DO

! Copy new fields into current fields
DO J=1,N+1
  DO I=1,M+1
    Ufld(I,J) = UNEW(I,J)
    Vfld(I,J) = VNEW(I,J)
    Pfld(I,J) = PNEW(I,J)
  END DO
END DO
\end{verbatim}
\caption{Example of the coding structure that required manual
  in-lining of the kernel body (as indicated by commented-out lines)
  to retrieve performance of original shallow with the Intel
  compiler.}
\label{FIG_time_smooth_code}
\end{figure}
